\chapter{Dynamic Equations}

In this chapter the forces and moments exerted on the aircraft will be presented. The main dynamic components are gravity, aerodynamic reactions and propulsion (also referred to as thrust).
\begin{equation}
	\bm{F}_b = \bm{F}_g + \bm{F}_a + \bm{F}_t
\end{equation}

Naturally, we consider that gravity does not exert moments on the rigid body aircraft.

\begin{equation}
	\bm{T}_b = \bm{T}_{tot,a} + \bm{T}_{tot,t}
\end{equation}

Before proceeding to any definitions, it is worth mentioning the control input variables. In normal airplane configurations, there are four control dimensions which the pilot can affect. These correspond to three sets of control surfaces (aileron, elevator and rudder) and the throttle command. The notation for these control variables is $\delta_a, \delta_a, \delta_t, \delta_r$. Essentially, these are the input variables to the aircraft model.$\delta_a, \delta_e,$ and $\delta_r$ are usually expressed in degrees of deflection from the neutral (zero) angle and $\delta_t$ is normalized in the (0,1) range.

Conventions for the positive direction of each control surface are not always consistent. In this text, the proposed positive direction of a control surface is the one which results in a positive moment around each primary axis (aileron - roll right, elevator - pitch up, rudder - yaw right).

%% Gravity section
\section{Gravity}

Initially, we can assume a constant value for the acceleration of gravity, $g_0=9.805416m/s^2$ \cite[p.~8]{USStdAtm76}. The magnitude of the acceleration of gravity normally varies with altitude and geodetic coordinates, but a constant value is a valid approach for low altitude flight; unmodeled components are negligible to an adequate accuracy. If a more detailed gravity model is required (for example, in flights with vertical variance in the scale of kilometers), the reader can turn to Section \ref{sec:earth_model}. The orientation of the gravity vector is always vertical, towards the center of the Earth, thus in the z-direction in the NED frame.

\begin{equation} \label{eq:gravForce}
	\bm{F}_{g} = \bm{R}_b
	\begin{bmatrix}
		0\\ 0 \\mg
	\end{bmatrix}
\end{equation}
\begin{IEEEeqnarray}{rCl}
	{F}_{gx} &= &-\sin \theta ~m g \IEEEyessubnumber\\
	{F}_{gy} &= & \sin \phi \cos \theta ~m g  \IEEEyessubnumber\\
	{F}_{gz} &= & \cos \phi \cos \theta ~m g \IEEEyessubnumber
\end{IEEEeqnarray}

\begin{lstlisting}[style=C-style]
	F_g_x = -sin(theta)*m*g
	F_g_y = sin(phi)*cos(theta)*m*g
	F_g_z = cos(phi)*cos(theta)*m*g
\end{lstlisting}

\begin{equation}\label{eq:gravity}
	g = g_0
\end{equation}

%% Aerodynamics section 
\input{aerodynamics}

%% Propulsion section
\section{Propulsion}

The devices which produce propulsion force in an aircraft can be termed \textit{thrusters}, in general. In the case of fixed-wing aircraft, most of the times thrusters take the form of a single propeller driven by a motor, usually acting upon the longitudinal axis of the aircraft, placed either in the front or the back of the fuselage. This discussion mainly draws from \cite[p.~127]{Allerton2009} but extends its contents from other sources to adapt to an electric power plant.
Thruster placement does not have an impact on the set of equations which form the thruster model, even in the case of twin-propeller configurations, mounted on the wings.

The propulsion force, expressed in the body-frame is declared as
\begin{equation} \label{eq:thrustForce}
	\bm{F}_t = \begin{bmatrix}
		F_{tx} \\ F_{ty} \\ F_{tz}
	\end{bmatrix}
\end{equation}


Usually, thrust is considered to be produced only along the body x-axis, however, this approach may lack realism: it is not uncommon for the propeller to be mounted slightly tilted right and downwards, in order to minimize the counter-moment it produces and which affects the aircraft stability negatively.

However, under the simplification that the propeller is aligned with the body frame x-axis, the following is also true:
\begin{IEEEeqnarray}{rCl} 
	F_{ty} &= & 0 \IEEEyessubnumber \\
	F_{tz} &= & 0 \IEEEyessubnumber
\end{IEEEeqnarray}

\begin{lstlisting}[style=C-style]
	F_t_y = 0
	F_t_z = 0
\end{lstlisting}

\todo[inline]{Write equations for thruster of random position and orientation}

Naturally, the motor and propeller produce moments. Similarly to the aerodynamic moment analysis, we also have to take into account the displacement of the trust vector from the x-axis of the body frame.
\begin{equation} \label{eq:thrustTorque}
	\bm{T_t} = \begin{bmatrix}
		T_{tx} \\ T_{ty} \\ T_{tz}
	\end{bmatrix}
	 + (\bm{p}_{prop}-\bm{p}_{CG})\times \bm{F}_p
\end{equation}
\begin{IEEEeqnarray}{rCl}
	T_{tx,tot} &= & T_{tx} -dz F_{ty} + dy F_{tz}\IEEEyessubnumber\\
	T_{ty,tot} &= & T_{ty} +dz F_{tx} - dx F_{tz}\IEEEyessubnumber\\
	T_{tz,tot} &= & T_{tz} -dy F_{tx} + dx F_{ty}\IEEEyessubnumber
\end{IEEEeqnarray}

\begin{lstlisting}[style=C-style]
dx = p_prop_x - p_cm_x
dy = p_prop_y - p_cm_y
dz = p_prop_z - p_cm_z
T_ttot_x = T_t_x - dz*F_t_y + dy*F_t_z
T_ttot_y = T_t_y + dz*F_t_x - dx*F_t_z
T_ttot_z = T_t_z - dy*F_t_z + dx*F_t_y
\end{lstlisting}

We opt to consider in our model only the aforementioned counter-moment produced in the x-axis by the propeller and motor, and not gyroscopic effects caused by the spinning propeller disc:
\begin{IEEEeqnarray}{rCl} 
	T_{ty} &= & 0 \IEEEyessubnumber \\
	T_{tz} &= & 0 \IEEEyessubnumber
\end{IEEEeqnarray}

\begin{lstlisting}[style=C-style]
	T_t_y = 0
	T_t_z = 0
\end{lstlisting}

With the aforementioned assumptions, the moment equations can be simplified into
\begin{IEEEeqnarray}{rCl}
	T_{ax,tot} &= & T_{tx}\IEEEyessubnumber\\
	T_{ay,tot} &= & dz F_{ax}\IEEEyessubnumber\\
	T_{az,tot} &= & -dy F_{ax}\IEEEyessubnumber
\end{IEEEeqnarray}

\begin{lstlisting}[style=C-style]
	T_ttot_x = T_t_x
	T_ttot_y = dz*F_a_x
	T_ttot_z = -dy*F_a_x
\end{lstlisting}

In the next subsection, a simplified model of a lumped thruster will be presented, as presented in \cite{Beard2012}. Afterwards, more detailed models will be presented, where the propeller model will be decoupled from the motor model. This enables the possibility to take advantage of the large amount of resources on propeller data and use them independently of the motor selection. The correct selection of the proper propeller for a given motor and vice versa can also be studied with this approach. Models for both electric and internal combustion motors will be presented.

%% Thrust control model
\input{motor_none}

%% Propeller model
\input{propeller}

%% Simplistic motor model
\input{motor_simple}

%% Electric motor model
\input{motor_electric}

%% IC engine model
\input{motor_ic}

